Hippocampal replay of experience at real-world speeds

### 2020-12-12 Article.tex
Reformat to eLife format

\documentclass[times, twoside]{zHenriquesLab-StyleBioRxiv}
\usepackage{blindtext}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{blkarray, bigstrut}
\usepackage{bbold}
\usepackage{amsmath}
\DeclareMathOperator*{\argmax}{arg\,max}

% Please give the surname of the lead author for the running footer
\leadauthor{Denovellis}

@@

\title{Hippocampal replay of experience at real-world speeds}
\shorttitle{Replay Dynamics}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[1]{Eric L. Denovellis}
\author[2, 3]{Anna K. Gillespie}
\author[2, 3]{Michael E. Coulter}
\author[4]{Marielena Sosa}
\author[5]{Jason E. Chung}
\author[6]{Uri T. Eden}
\author[1, 2, 3, \Letter]{Loren M. Frank}

@@

@@

%TC:break Abstract
%the command above serves to have a word count for the abstract
\begin{abstract}
Representations of past and possible future experiences play a critical role in memory and decision-making processes. The hippocampus expresses these types of representations during sharp-wave ripple (SWR) events, and previous work identified a minority of SWRs that contain “replay” of spatial trajectories at \texttildelow20x real-world speeds. Efforts to understand replay typically make multiple assumptions about which events to examine and what sorts of representations constitute replay. We therefore lack a clear understanding of both the prevalence and the range of representational dynamics associated with replay. Here we develop a state space model that uses a combination of movement dynamics of different speeds to capture the spatial content and time evolution of replay during SWRs. Using this model, we find that the large majority of replay events contain spatially coherent, interpretable content. Furthermore, most events progress at real-world, rather than accelerated, movement speeds, consistent with actual experiences. 
\end {abstract}
%TC:break main
%the command above serves to have a word count for the abstract

\begin{keywords}
Hippocampus | Replay | State Space | Dynamics
\end{keywords}

\begin{corrauthor}
%\texttt{loren{@}phy.ucsf.edu}
loren\at phy.ucsf.edu
\end{corrauthor}

@@

The brain has the remarkable ability to retrieve representations of past events and generate representations of possible future events. These "generative" \cite{KayConstantSubsecondCycling2020} mental events may not occur in the same time span as actual events: memories can be retrieved and used in less time than was required for the original experience, and mental simulations can span a series of actions more quickly than would be required to perform those actions. Thus, a potential neural substrate for retrieval and mental simulation would be expected to support rapid reinstatement of patterns of brain activity associated with past or potential future experience.

Hippocampal "replay" events are well suited to that role \cite{CarrHippocampalreplayawake2011}. As animals move through space, neurons in the hippocampus preferentially fire at specific locations in an environment, and thus sets of cells fire in sequence as the animal moves through a series of locations. When the animal is asleep or immobile, hippocampal cells can be reactivated during a "sharp-wave ripple" (SWR) event. A subset of SWRs contain sequential firing similar to that seen during a previous experience, and are thus thought to "replay" these previous experiences. Importantly, previous work has reported that these sequential firing events proceed at an average speed of \texttildelow10 meters per second, about 20x faster than the animal's usual movement speed \cite{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}.

While the existence of these sequential events is well established, the current consensus is that only a minority (\texttildelow10-45\%) of hippocampal SWRs contain statistically identifiable, sequential replay \cite{MichonPostlearningHippocampalReplay2019, KaeferReplayBehavioralSequences2020, ShinDynamicsAwakeHippocampalPrefrontal2019, DavidsonHippocampalReplayExtended2009}. One might therefore conclude that the retrieval of stored representation during SWRs is limited to sequences of locations, and that all such retrieval events are sequential in nature. Our subjective experience is that we can retrieve memories of locations without having to mentally traverse a long distance. If the rodent hippocampus is capable of that sort of retrieval, then one might expect to see SWRs where the neural activity corresponds not to a very rapid trajectory through space but instead to either a stable pattern associated with a single location or perhaps a pattern more similar to one seen during a real experience. Importantly, this sort of activity could still be useful, as it would take place during the short duration of an SWR (\texttildelow100 ms) and could thus support rapid retrieval and use of representations.

Interestingly, there is evidence for SWRs where the spiking corresponds to a single location. Specifically, some SWRs contain spiking of neurons associated with single locations where animals are immobile \cite{JaiDistincthippocampalcorticalmemory2017}, but another report suggested that events that replay a single location are only seen in young animals \cite{StellaHippocampalReactivationRandom2019}. Thus, while it would seem useful to replay memories associated with single locations and perhaps even experiences occurring at real-world speeds, the prevalence of that type of replay in awake animals remains unclear.

Our uncertainty stems in part from the dominant approaches used to identify the content of replay events. These approaches typically involve multiple steps and multiple assumptions about the nature of replay, which are most commonly characterized using the standard "Bayesian" decoder. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on a threshold for event size chosen by the experimenter \cite{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009, StellaHippocampalReactivationRandom2019}. A decoding algorithm is then applied to the spikes within these events, yielding a set of position probability distributions for each time bin. Current approaches use either overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approach for detecting sequential replay involves fitting a line to the resulting set of probability distributions, which relies on the assumption that the representations progress at a constant, non-zero speed \cite{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009}. A statistical test is then used to determine whether the fit is better than the fit to shuffled versions of the data.

While the standard decoder approach identifies constant speed events, it does not consider events that are rejected by the statistical test, and the use of a fixed size temporal bins acts as a boxcar smoother that limits the potential movement speeds of the representation. The linear fit is also problematic: it has the potential to reject or mischaracterize real events that do not move at constant speeds. Approaches that focus on the order of cell activity within each event \cite{LeeMemorySequentialExperience2002, GuptaHippocampalReplayNot2010} allow for a relaxation of that linear assumption, but replace it with a loss of statistical power due to either ignoring all but the first spike from each cell or using an arbitrarily smoothed version of the spike train. These approaches also do not provide information about the dynamics of the underlying spatial representation. Moreover, these approaches exclude events that  have stationary representations of a single location \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}.

Recognizing the problems with the linear fit, several studies have moved away from the constant velocity assumption, using an estimated represented location at each time bin and connecting each with a line. For example, using this approach, Pfeiffer and Foster \cite{PfeifferAutoassociativedynamicsgeneration2015} found that awake replays can alternate between representing a single location and sequential spatial trajectories. On the other hand, Stella et al. \cite{StellaHippocampalReactivationRandom2019} reported that replays during sleep are spatially continuous and follow Brownian diffusion dynamics. Both methods still used large time bins and neither took into account the uncertainty of the decoded estimates, however, making it hard to identify the source of the different conclusions.

@@

We therefore developed a state space model of generative activity that achieves all of those goals. State space models are a well-understood, well-known statistical solution to the problems described above. By mathematically modeling the relationship between the data and latent dynamics, state space models make the assumptions of the model explicit and interpretable. Our model goes beyond previous approaches \cite{MaboudiUncoveringtemporalstructure2018, DengRapidclassificationhippocampal2016} by characterizing represented trajectories as a mixture of three underlying patterns of movement dynamics: stationary trajectories, continuous trajectories that can progress at many times the typical speed of the animal, and spatially fragmented trajectories. We show how this model can take advantage of clusterless decoding---which relates multiunit spike waveform features to position without spike sorting---giving us more information about the population spiking activity. We apply this model to spiking data during SWR events from 10 rats, enabling a direct comparison to previous work.

@@

\section*{Results}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v6}
\caption{The model can capture different sequence dynamics on simulated data. \textbf{(A)} We construct a firing sequence of 19 simulated place cells that exhibits three different movement dynamics. For the first 60 ms, one cell fires repeatedly, representing one stationary location. For next 190 ms, the cells fire in sequence, representing a rapid continuous trajectory along the virtual track. Finally, for the last 30 ms, cells fire randomly, out of spatial order, representing an fragmented spatial sequence. \textbf{(B)} Like the standard decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} The prediction from the neural data is then combined with an explicit model of each movement dynamic, which determines how latent position can change based on the position in the previous time step. We show the probability of the next position bin for each movement dynamic model (color scale). Zero here represents the previous position. \textbf{(D)} The probability of remaining in a particular movement dynamic versus switching to another dynamic is modeled as having a high probability of remaining in a particular dynamic with a small probability of switching to one of the other dynamics at each time step. \textbf{(E)} The model uses the components in A-D over all time to decode the joint posterior probability of latent position and dynamic. This can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic over time. The shaded colors indicate the category of the speed of the trajectory at that time, which is determined from the probability. Marginalizing the posterior across dynamics also provides an estimate of latent position over time (right panel). \textbf{(F)} The probability of each dynamic depends heavily on the speed of the trajectory, as we show using a range of simulated spiking sequences each moving at a constant speed. Each dot corresponds to the average probability of that dynamic for a given constant speed sequence. We use a 0.8 threshold (dotted line) to classify each sequence based on the dynamic or dynamics which contribute maximally to the posterior (shaded colors).
}
\label{1}
\end{figure*}
\subsection*{Overview of the model}
We begin with an application of the state space model to simulated data, both to validate the model and to provide intuition (Figure \ref{1}). We simulate 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then construct a 280 ms spiking sequence (Figure \ref{1}A) and apply our model to the sequence. For the first 60 ms of this sequence, a single place cell fires repeatedly, resulting in the extended representation of a single location. For the second 190 ms of the sequence, the cells fire in sequential spatial order, representing a fast moving trajectory across the virtual linear track. For the last 30 ms of the sequence, the cells fire in an incoherent spatial order. These three firing patterns represent three different types of movement dynamics that could be expressed during SWRs, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize SWRs in terms of a mixture of these three dynamics at every time point.

Decoding the spiking sequence requires specifying two elements: the data model---how the spikes relate to position---and the movement dynamic models---how the position can change over time in each movement dynamic. For the data model, our decoder is the same as the standard ("Bayesian") decoder \cite{DavidsonHippocampalReplayExtended2009, PfeifferAutoassociativedynamicsgeneration2015, StellaHippocampalReactivationRandom2019}. We compute an estimate of how each cell's firing rate varies over position during movement (i.e. the place field, Figure \ref{1}B). This is used during decoding to compute the Poisson likelihood of position over time given the spiking sequence of interest. In contrast to the standard decoding approaches, we can use small time bins (in our case 2 ms vs. 20 ms or more) because we are able to take advantage of the prior placed on the dynamics by the state space model. This allows us to detect changes on smaller time scales than would be possible with the standard decoder.

Next, we specify movement dynamic models for how latent position---the "mental" position of the animal represented by the cells---evolves over time for a given dynamic (Figure \ref{1}C). We do this by defining a state transition matrix that defines how the latent position can move from the previous time step (in our case, 2 ms). Previous findings suggest that replay may exhibit at least three distinct types of movement dynamics: stationary \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, continuous \cite{DavidsonHippocampalReplayExtended2009}, and fragmented, which could correspond to both extended spatially incoherent representations and representations that jump from one position to another in a short time period \cite{PfeifferAutoassociativedynamicsgeneration2015}. We therefore define state transition models to capture each of these movement dynamics.

In the stationary movement dynamic, the latent position does not change between time steps. The state transition matrix can thus be defined as an identity matrix, which predicts the next position will be the same as the last position (subject to the resolution of the position bin). In the continuous movement dynamic, the latent position is most likely to be "spatially close" to the position in the previous time step, so we use a Gaussian random walk state transition matrix. This means that, for a given latent position, the probability of moving to another position is modeled by a Gaussian centered at that position and "spatially close" is defined by the variance of the Gaussian. In our case, since replay has been shown to move at speeds much faster than the animal's speed \cite{DavidsonHippocampalReplayExtended2009, PfeifferAutoassociativedynamicsgeneration2015}, we set the variance of the Gaussian to 6.0 cm. This ensures that with a 2 ms time step, the latent position is 95\% likely to be within 4.90 cm of the previous latent position (or equivalently, this means that latent speeds from 0 to \texttildelow25 m/s are most likely).  Finally, in the fragmented movement dynamic, the latent position can move to any available position instantaneously. We model this using a uniform state transition matrix, which makes transitions to each position equally likely.

Finally, we specify how likely each movement dynamic is to persist in time versus change to another dynamic (Figure \ref{1}D) via a state transition matrix between dynamics. In order to be conservative with respect to switching between dynamics, we assume that each movement dynamic is likely to dominate for \texttildelow100 ms, which is approximately the duration of a SWR event. There is, however, a small probability of switching to one of the other movement dynamics. Accordingly, we set the probability of staying in a dynamic to 0.98 for each 2 ms time step, which corresponds to an expected duration of 100 ms for staying in a particular dynamic (the Markov assumption of the model means that the probability of staying in a dynamic follows a geometric distribution). Importantly, the data drives the estimated dynamic, so even if the probability of a particular movement dynamic is high, data providing clear evidence for a change in dynamic can, within the time frame of \texttildelow10 ms, drive a change to a different estimated dynamic, as in our simulated example. In line with this, note that below we show that our results are relatively insensitive to the value of this parameter.

Once we have specified the data and the movement dynamic models, we have fully specified the state space model. We use acausal decoding, meaning that we use all information from both past and future spikes, to estimate the joint posterior probability of position and dynamic (see Methods). With this, we can summarize the resulting posterior probability with two quantities: the probability of each movement dynamic over time (by integrating out position) and the probability of latent position over time, irrespective of movement dynamic (by summing over the movement dynamics; Figure \ref{1}E, left and right plot respectively).

An examination of the two summaries shown in Figure \ref{1}E reveals that that the model successfully captures the dynamics of the population spiking activity in Figure \ref{1}A. The stable firing of the one active neuron indicates a stationary representation, and accordingly, the probability of the stationary movement dynamic is high at the beginning of the simulated replay. A change in the data then drives a rapid transition to the continuous movement dynamic, reflecting the trajectory-like spatially sequential spiking from the population of cells. Subsequently, as the population activity becomes spatially incoherent, the fragmented movement dynamic dominates for the last 30 ms of the simulated event.

Importantly, this approach allows the model to to capture a wide range of movement speeds for the latent position. The model is defined in terms of a mixture of movement dynamics, as summarized by the probability of each movement dynamic, and which dynamic or dynamics are dominant at a given moment is related to the movement of the underlying position representation. To demonstrate this, we applied the model to 10,000 simulated trajectories, each trajectory proceeding at a constant speed (Figure \ref{1}F) from 1 cm/s to 10,000 m/s. From this, we can see that not only are there regions of speed that correspond to each of our three movement dynamics being highly probable (where we define highly probable to be greater than or equal to 0.8 probability), but there are also intermediate speeds where two of the dynamics exhibit relatively high probability; and where the sum of two of the dynamics' probabilities exceeds 0.8. In this manuscript, we will refer to these intermediate speeds as mixture dynamics. For example, when the stationary dynamic has a probability of 0.6 and the continuous has a probability of 0.4, we call this a stationary-continuous mixture (light blue, Figure \ref{1}F) and this indicates that the trajectory is moving slowly. Correspondingly, if the continuous dynamic has a probability of 0.5 and the fragmented dynamic has a probability of 0.4, then we would call this a fragmented-continuous-mixture and this indicates the trajectory is moving very quickly, but not as quickly as the fragmented dynamic dictates. In summary, we can characterize the speed or set of speeds in a SWR based on the probability of each of the three movement dynamics over time. We further classify the probability of each movement dynamic as being part of one of five speed categories: stationary, stationary-continuous-mixtures, continuous, fragmented-continuous mixtures, and fragmented.

@@

Finally, as mentioned above, we wanted to test the robustness of the model to the choice of probability of staying in a dynamic, because our choice of 0.98 or an expected duration of 100 ms is only based on the expected duration of a SWR. To investigate this we decoded the spiking sequence in Figure \ref{1}A with different probabilities of staying in the same dynamic versus switching to another dynamic (Figure \ref{fig:Figure1-Figure supplement 1}). We found that for a large range of plausible probabilities of staying in one of the dynamics (between 0.96 and 0.993 which corresponds to an expected duration between 25 and 150 ms), the model still correctly identified the sequence dynamics with high probability (Figure \ref{fig:Figure1-Figure supplement 1}A), demonstrating that data itself is the most influential element in the model. Furthermore, the probability of latent position remains relatively consistent across the range of these probabilities as well (Figure \ref{fig:Figure1-Figure supplement 1}B).

>>>>

\documentclass[9pt,lineno]{elife}
\usepackage{blindtext}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{blkarray, bigstrut}
\usepackage{bbold}
\usepackage{amsmath}
\DeclareMathOperator*{\argmax}{arg\,max}

@@

\title{Hippocampal replay of experience at real-world speeds}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[1]{Eric L. Denovellis}
\author[2, 3]{Anna K. Gillespie}
\author[2, 3]{Michael E. Coulter}
\author[4]{Marielena Sosa}
\author[5]{Jason E. Chung}
\author[6]{Uri T. Eden}
\author[1, 2, 3]{Loren M. Frank}

@@

\corr{loren@phy.ucsf.edu}{LMF}

@@

\begin{abstract}
Representations of past and possible future experiences play a critical role in memory and decision-making processes. The hippocampus expresses these types of representations during sharp-wave ripple (SWR) events, and previous work identified a minority of SWRs that contain “replay” of spatial trajectories at \texttildelow20x real-world speeds. Efforts to understand replay typically make multiple assumptions about which events to examine and what sorts of representations constitute replay. We therefore lack a clear understanding of both the prevalence and the range of representational dynamics associated with replay. Here we develop a state space model that uses a combination of movement dynamics of different speeds to capture the spatial content and time evolution of replay during SWRs. Using this model, we find that the large majority of replay events contain spatially coherent, interpretable content. Furthermore, most events progress at real-world, rather than accelerated, movement speeds, consistent with actual experiences. 
\end {abstract}

@@

The brain has the remarkable ability to retrieve representations of past events and generate representations of possible future events. These "generative" \citep{KayConstantSubsecondCycling2020} mental events may not occur in the same time span as actual events: memories can be retrieved and used in less time than was required for the original experience, and mental simulations can span a series of actions more quickly than would be required to perform those actions. Thus, a potential neural substrate for retrieval and mental simulation would be expected to support rapid reinstatement of patterns of brain activity associated with past or potential future experience.

Hippocampal "replay" events are well suited to that role \citep{CarrHippocampalreplayawake2011}. As animals move through space, neurons in the hippocampus preferentially fire at specific locations in an environment, and thus sets of cells fire in sequence as the animal moves through a series of locations. When the animal is asleep or immobile, hippocampal cells can be reactivated during a "sharp-wave ripple" (SWR) event. A subset of SWRs contain sequential firing similar to that seen during a previous experience, and are thus thought to "replay" these previous experiences. Importantly, previous work has reported that these sequential firing events proceed at an average speed of \texttildelow10 meters per second, about 20x faster than the animal's usual movement speed \citep{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}.

While the existence of these sequential events is well established, the current consensus is that only a minority (\texttildelow10-45\%) of hippocampal SWRs contain statistically identifiable, sequential replay \citep{MichonPostlearningHippocampalReplay2019, KaeferReplayBehavioralSequences2020, ShinDynamicsAwakeHippocampalPrefrontal2019, DavidsonHippocampalReplayExtended2009}. One might therefore conclude that the retrieval of stored representation during SWRs is limited to sequences of locations, and that all such retrieval events are sequential in nature. Our subjective experience is that we can retrieve memories of locations without having to mentally traverse a long distance. If the rodent hippocampus is capable of that sort of retrieval, then one might expect to see SWRs where the neural activity corresponds not to a very rapid trajectory through space but instead to either a stable pattern associated with a single location or perhaps a pattern more similar to one seen during a real experience. Importantly, this sort of activity could still be useful, as it would take place during the short duration of an SWR (\texttildelow100 ms) and could thus support rapid retrieval and use of representations.

Interestingly, there is evidence for SWRs where the spiking corresponds to a single location. Specifically, some SWRs contain spiking of neurons associated with single locations where animals are immobile \citep{JaiDistincthippocampalcorticalmemory2017}, but another report suggested that events that replay a single location are only seen in young animals \citep{StellaHippocampalReactivationRandom2019}. Thus, while it would seem useful to replay memories associated with single locations and perhaps even experiences occurring at real-world speeds, the prevalence of that type of replay in awake animals remains unclear.

Our uncertainty stems in part from the dominant approaches used to identify the content of replay events. These approaches typically involve multiple steps and multiple assumptions about the nature of replay, which are most commonly characterized using the standard "Bayesian" decoder. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on a threshold for event size chosen by the experimenter \citep{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009, StellaHippocampalReactivationRandom2019}. A decoding algorithm is then applied to the spikes within these events, yielding a set of position probability distributions for each time bin. Current approaches use either overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approach for detecting sequential replay involves fitting a line to the resulting set of probability distributions, which relies on the assumption that the representations progress at a constant, non-zero speed \citep{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009}. A statistical test is then used to determine whether the fit is better than the fit to shuffled versions of the data.

While the standard decoder approach identifies constant speed events, it does not consider events that are rejected by the statistical test, and the use of a fixed size temporal bins acts as a boxcar smoother that limits the potential movement speeds of the representation. The linear fit is also problematic: it has the potential to reject or mischaracterize real events that do not move at constant speeds. Approaches that focus on the order of cell activity within each event \citep{LeeMemorySequentialExperience2002, GuptaHippocampalReplayNot2010} allow for a relaxation of that linear assumption, but replace it with a loss of statistical power due to either ignoring all but the first spike from each cell or using an arbitrarily smoothed version of the spike train. These approaches also do not provide information about the dynamics of the underlying spatial representation. Moreover, these approaches exclude events that  have stationary representations of a single location \citep{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}.

Recognizing the problems with the linear fit, several studies have moved away from the constant velocity assumption, using an estimated represented location at each time bin and connecting each with a line. For example, using this approach, \cite{PfeifferAutoassociativedynamicsgeneration2015} found that awake replays can alternate between representing a single location and sequential spatial trajectories. On the other hand, \cite{StellaHippocampalReactivationRandom2019} reported that replays during sleep are spatially continuous and follow Brownian diffusion dynamics. Both methods still used large time bins and neither took into account the uncertainty of the decoded estimates, however, making it hard to identify the source of the different conclusions.

@@

We therefore developed a state space model of generative activity that achieves all of those goals. State space models are a well-understood, well-known statistical solution to the problems described above. By mathematically modeling the relationship between the data and latent dynamics, state space models make the assumptions of the model explicit and interpretable. Our model goes beyond previous approaches \citep{MaboudiUncoveringtemporalstructure2018, DengRapidclassificationhippocampal2016} by characterizing represented trajectories as a mixture of three underlying patterns of movement dynamics: stationary trajectories, continuous trajectories that can progress at many times the typical speed of the animal, and spatially fragmented trajectories. We show how this model can take advantage of clusterless decoding---which relates multiunit spike waveform features to position without spike sorting---giving us more information about the population spiking activity. We apply this model to spiking data during SWR events from 10 rats, enabling a direct comparison to previous work.

@@

\section*{Results}

% Figure 1
\begin{figure}
\includegraphics[width=0.75\linewidth]{figures/Figure1/Figure1_v6}
\caption{The model can capture different sequence dynamics on simulated data. \textbf{(A)} We construct a firing sequence of 19 simulated place cells that exhibits three different movement dynamics. For the first 60 ms, one cell fires repeatedly, representing one stationary location. For next 190 ms, the cells fire in sequence, representing a rapid continuous trajectory along the virtual track. Finally, for the last 30 ms, cells fire randomly, out of spatial order, representing an fragmented spatial sequence. \textbf{(B)} Like the standard decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} The prediction from the neural data is then combined with an explicit model of each movement dynamic, which determines how latent position can change based on the position in the previous time step. We show the probability of the next position bin for each movement dynamic model (color scale). Zero here represents the previous position. \textbf{(D)} The probability of remaining in a particular movement dynamic versus switching to another dynamic is modeled as having a high probability of remaining in a particular dynamic with a small probability of switching to one of the other dynamics at each time step. \textbf{(E)} The model uses the components in A-D over all time to decode the joint posterior probability of latent position and dynamic. This can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic over time. The shaded colors indicate the category of the speed of the trajectory at that time, which is determined from the probability. Marginalizing the posterior across dynamics also provides an estimate of latent position over time (right panel). \textbf{(F)} The probability of each dynamic depends heavily on the speed of the trajectory, as we show using a range of simulated spiking sequences each moving at a constant speed. Each dot corresponds to the average probability of that dynamic for a given constant speed sequence. We use a 0.8 threshold (dotted line) to classify each sequence based on the dynamic or dynamics which contribute maximally to the posterior (shaded colors).
}
\label{fig:1}
% Figure 1-Supplemental 1
\figsupp[The model is robust to change of the probability of persisting in the same dynamic for a wide range of plausible expected durations (25 ms to 150 ms)]{The model is robust to change of the probability of persisting in the same dynamic for a wide range of plausible expected durations (25 ms to 150 ms). \textbf{(A)} Each plot shows the probability of each dynamic on simulated data example from \FIG{1} with a different diagonal value---which governs the probability of remaining in that dynamic. The corresponding expected duration of staying in the dynamic is listed as duration. The off-diagonal values---the probability of switching to one of the other dynamics---are set to be equally likely with the remainder of the probability, as in \FIG{1}D. The diagonal increases from left to right, top to bottom, until the case where the diagonal is one and the off-diagonal is zero---i.e. the case where there is no probability of switching to another dynamic. Shaded regions correspond to the classification as in \FIG{1}. \textbf{(B)} The probability of position over time for each diagonal value. Conventions the same as in (A).}{\includegraphics[width=1.0\linewidth]{figures/Figure1-supplemental1/Figure1_v3_supplemental1}}\label{figsupp:fig1sup1}
\end{figure}

\subsection*{Overview of the model}
We begin with an application of the state space model to simulated data, both to validate the model and to provide intuition (\FIG{1}). We simulate 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then construct a 280 ms spiking sequence (\FIG{1}A) and apply our model to the sequence. For the first 60 ms of this sequence, a single place cell fires repeatedly, resulting in the extended representation of a single location. For the second 190 ms of the sequence, the cells fire in sequential spatial order, representing a fast moving trajectory across the virtual linear track. For the last 30 ms of the sequence, the cells fire in an incoherent spatial order. These three firing patterns represent three different types of movement dynamics that could be expressed during SWRs, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize SWRs in terms of a mixture of these three dynamics at every time point.

Decoding the spiking sequence requires specifying two elements: the data model---how the spikes relate to position---and the movement dynamic models---how the position can change over time in each movement dynamic. For the data model, our decoder is the same as the standard ("Bayesian") decoder \citep{DavidsonHippocampalReplayExtended2009, PfeifferAutoassociativedynamicsgeneration2015, StellaHippocampalReactivationRandom2019}. We compute an estimate of how each cell's firing rate varies over position during movement (i.e. the place field, \FIG{1}B). This is used during decoding to compute the Poisson likelihood of position over time given the spiking sequence of interest. In contrast to the standard decoding approaches, we can use small time bins (in our case 2 ms vs. 20 ms or more) because we are able to take advantage of the prior placed on the dynamics by the state space model. This allows us to detect changes on smaller time scales than would be possible with the standard decoder.

Next, we specify movement dynamic models for how latent position---the "mental" position of the animal represented by the cells---evolves over time for a given dynamic (\FIG{1}C). We do this by defining a state transition matrix that defines how the latent position can move from the previous time step (in our case, 2 ms). Previous findings suggest that replay may exhibit at least three distinct types of movement dynamics: stationary \citep{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, continuous \citep{DavidsonHippocampalReplayExtended2009}, and fragmented, which could correspond to both extended spatially incoherent representations and representations that jump from one position to another in a short time period \citep{PfeifferAutoassociativedynamicsgeneration2015}. We therefore define state transition models to capture each of these movement dynamics.

In the stationary movement dynamic, the latent position does not change between time steps. The state transition matrix can thus be defined as an identity matrix, which predicts the next position will be the same as the last position (subject to the resolution of the position bin). In the continuous movement dynamic, the latent position is most likely to be "spatially close" to the position in the previous time step, so we use a Gaussian random walk state transition matrix. This means that, for a given latent position, the probability of moving to another position is modeled by a Gaussian centered at that position and "spatially close" is defined by the variance of the Gaussian. In our case, since replay has been shown to move at speeds much faster than the animal's speed \citep{DavidsonHippocampalReplayExtended2009, PfeifferAutoassociativedynamicsgeneration2015}, we set the variance of the Gaussian to 6.0 cm. This ensures that with a 2 ms time step, the latent position is 95\% likely to be within 4.90 cm of the previous latent position (or equivalently, this means that latent speeds from 0 to \texttildelow25 m/s are most likely).  Finally, in the fragmented movement dynamic, the latent position can move to any available position instantaneously. We model this using a uniform state transition matrix, which makes transitions to each position equally likely.

Finally, we specify how likely each movement dynamic is to persist in time versus change to another dynamic (\FIG{1}D) via a state transition matrix between dynamics. In order to be conservative with respect to switching between dynamics, we assume that each movement dynamic is likely to dominate for \texttildelow100 ms, which is approximately the duration of a SWR event. There is, however, a small probability of switching to one of the other movement dynamics. Accordingly, we set the probability of staying in a dynamic to 0.98 for each 2 ms time step, which corresponds to an expected duration of 100 ms for staying in a particular dynamic (the Markov assumption of the model means that the probability of staying in a dynamic follows a geometric distribution). Importantly, the data drives the estimated dynamic, so even if the probability of a particular movement dynamic is high, data providing clear evidence for a change in dynamic can, within the time frame of \texttildelow10 ms, drive a change to a different estimated dynamic, as in our simulated example. In line with this, note that below we show that our results are relatively insensitive to the value of this parameter.

Once we have specified the data and the movement dynamic models, we have fully specified the state space model. We use acausal decoding, meaning that we use all information from both past and future spikes, to estimate the joint posterior probability of position and dynamic (see Methods). With this, we can summarize the resulting posterior probability with two quantities: the probability of each movement dynamic over time (by integrating out position) and the probability of latent position over time, irrespective of movement dynamic (by summing over the movement dynamics; \FIG{1}E, left and right plot respectively).

An examination of the two summaries shown in \FIG{1}E reveals that that the model successfully captures the dynamics of the population spiking activity in \FIG{1}A. The stable firing of the one active neuron indicates a stationary representation, and accordingly, the probability of the stationary movement dynamic is high at the beginning of the simulated replay. A change in the data then drives a rapid transition to the continuous movement dynamic, reflecting the trajectory-like spatially sequential spiking from the population of cells. Subsequently, as the population activity becomes spatially incoherent, the fragmented movement dynamic dominates for the last 30 ms of the simulated event.

Importantly, this approach allows the model to to capture a wide range of movement speeds for the latent position. The model is defined in terms of a mixture of movement dynamics, as summarized by the probability of each movement dynamic, and which dynamic or dynamics are dominant at a given moment is related to the movement of the underlying position representation. To demonstrate this, we applied the model to 10,000 simulated trajectories, each trajectory proceeding at a constant speed (\FIG{1}F) from 1 cm/s to 10,000 m/s. From this, we can see that not only are there regions of speed that correspond to each of our three movement dynamics being highly probable (where we define highly probable to be greater than or equal to 0.8 probability), but there are also intermediate speeds where two of the dynamics exhibit relatively high probability; and where the sum of two of the dynamics' probabilities exceeds 0.8. In this manuscript, we will refer to these intermediate speeds as mixture dynamics. For example, when the stationary dynamic has a probability of 0.6 and the continuous has a probability of 0.4, we call this a stationary-continuous mixture (light blue, \FIG{1}F) and this indicates that the trajectory is moving slowly. Correspondingly, if the continuous dynamic has a probability of 0.5 and the fragmented dynamic has a probability of 0.4, then we would call this a fragmented-continuous-mixture and this indicates the trajectory is moving very quickly, but not as quickly as the fragmented dynamic dictates. In summary, we can characterize the speed or set of speeds in a SWR based on the probability of each of the three movement dynamics over time. We further classify the probability of each movement dynamic as being part of one of five speed categories: stationary, stationary-continuous-mixtures, continuous, fragmented-continuous mixtures, and fragmented.

@@

Finally, as mentioned above, we wanted to test the robustness of the model to the choice of probability of staying in a dynamic, because our choice of 0.98 or an expected duration of 100 ms is only based on the expected duration of a SWR. To investigate this we decoded the spiking sequence in \FIG{1}A with different probabilities of staying in the same dynamic versus switching to another dynamic (\FIGSUPP[1]{fig1sup1}). We found that for a large range of plausible probabilities of staying in one of the dynamics (between 0.96 and 0.993 which corresponds to an expected duration between 25 and 150 ms), the model still correctly identified the sequence dynamics with high probability (\FIGSUPP[1]{fig1sup1}A), demonstrating that data itself is the most influential element in the model. Furthermore, the probability of latent position remains relatively consistent across the range of these probabilities as well (\FIGSUPP[1]{fig1sup1}B).
